Data Analysis Methods for the Detection of the Epoch of Reionization

### 2017-11-21 PAPER.tex
Made jonnie edits until 2.1

To build our intuition into how a particular choice of weighting, namely inverse covariance weighting, can lead to signal loss, we use a toy model. We construct a simple dataset that contains 2-dimensional data (representing visibility data with $100$ time integrations and $20$ frequency channels). This model represents realistic dimensions of about an hour of PAPER data which might be used for a power spectrum analysis. For PAPER-64 we use $\sim$$8$ hours of data for our analysis, but here we scale it down for this toy model with no loss of generality. We create a mock bright foreground signal, $\textbf{x}_{FG}$, as a complex sinusoid with a fixed frequency and a smoothly varying offset in time, a simplistic but realistic representation of a single bright source. We also create a mock EoR signal, $\textbf{x}_{EoR}$, as a complex, Gaussian random signal. A more realistic EoR signal would have a sloped power spectrum in $\hat{\textbf{p}}$(k) (instead of flat, as in the case of white noise), which could be simulated by introducing frequency correlations into the mock EoR signal. However, in this paper we treat all $k$'s separately, so a simplistic white noise approximation can be used. Our combined data vector is then $\textbf{x} = \textbf{x}_{FG} + \textbf{x}_{EoR}$, to which we apply different weighting schemes. The three data vectors are shown in Figure \ref{fig:toy_sigloss1}.

@@

One choice for the weighting matrix $\textbf{R}$ is an inverse covariance matrix. This type of weighting is attractive for power spectrum analyses because it is an aggressive way to down-weight foregrounds and yields the smallest possible error bars on a measurement (\citealt{tegmark_et_al1997a}; \citealt{bond_et_al1998}). The covariance matrix $\textbf{C}$, which in our case describes covariances between frequency channels, can be estimated in a variety of ways. Given perfect foreground, instrumental, and EoR models, we could form $\textbf{C}$ in a way that accurately describes our measured data. However, if our foreground model is flawed for example, our estimate of $\textbf{C}$ would not be successful at down-weighting them in the data.

@@

\noindent assuming $\langle\textbf{x}\rangle = 0$, where $\langle \rangle$ denotes a finite average over time $t$. The inverse covariance matrix is therefore $\textbf{R} = \hat{\textbf{C}}^{-1}$.

First, we compute the power spectrum of $\textbf{x}$ using OQE formalism and a weighting matrix of $\hat{\textbf{C}}^{-1}$. The result is shown in green in the left plot of Figure \ref{fig:toy_sigloss3}. Also plotted in the figure are the unweighted ($\textbf{R} = \textbf{I}$) power spectrum of $\textbf{x}_{FG}$ (blue) and $\textbf{x}_{EoR}$ (red).

@@

As shown, our inverse covariance-weighted result successfully suppresses foregrounds. It is also evident that our result fails to recover the EoR signal --- it exhibits the correct shape, but the amplitude level is slightly low. This is evidence of signal loss. In order to understand the behavior of this result, we can closely study our covariance matrix, $\hat{\textbf{C}}$, which is shown in Figure \ref{fig:toy_sigloss12}.

If $\textbf{C}$ is computed from the data itself, it carries the risk of over-fitting information in the data and introducing a multiplicative bias to estimates of the signal. For a mathematical derivation of signal loss arising from a data-estimated covariance matrix, see Appendix \ref{sec:sigloss_appendix}. Here we will describe the origin of this signal loss intuitively.

@@

In general, the strongest modes of $\hat{\textbf{C}}$ (highest eigenvalues) are more trustworthy than the weakest modes, since that's where we expect bright foregrounds --- the most prominent shapes --- to be. If bright foregrounds dominate, these highest eigenvalue modes would exist in both $\textbf{C}$ and $\hat{\textbf{C}}$. This is demonstrated in the toy model by the successful suppression of the foreground mode, where our estimated covariance matrix identifies the sinusoid shape and assigns it the highest eigenvalue (the peak in Figure \ref{fig:toy_sigloss2}).

@@

Using what we've learned about the eigenspectrum, we can tweak it in a simple way to suppress foregrounds and yield zero signal loss. Namely, we can project out the zeroth eigenmode (i.e. zero out all eigenmodes except the first one), thereby down-weighting the foregrounds perfectly and nothing else. Hence, we are not using information from all the weaker eigenmodes which carry the risk of describing fluctuations that may not exist in the true covariance. Altering $\hat{\textbf{C}}$ as such is one example of a regularization method, in which we are changing $\hat{\textbf{C}}$ in a way that flattens its eigenspectrum, making it more similar to that of $\textbf{I}$. The resulting power spectrum estimate for this case is shown in the right plot of Figure \ref{fig:toy_sigloss3}. In this case we recover EoR, demonstrating that if we can model our foregrounds perfectly (and disentangle them from other modes), we can down-weight them without signal loss. There are several other ways to regularize $\hat{\textbf{C}}$, and we will discuss some in Section \ref{sec:otherweight}.

@@

A fringe-rate filter is an analysis technique designed to maximize sensitivity by integrating in time (\citealt{parsons_et_al2016}). Rather than a traditional box-car average, a time domain filter can be designed to up-weight temporal modes relevant to things that move as expected for the sidereal motion on the sky, while down-weighting modes which are noise-like.

Because fringe-rate filtering is analogous to averaging in time, it comes at the cost of reducing the total number of independent samples in the data. To mimic this filter, we average every four time integrations of our toy model dataset together, yielding $25$ independent samples in time (Figure \ref{fig:toy_sigloss5}, left). We choose these numbers so that the total number of independent samples is similar to the number of frequency channels --- therefore, our matrices will be full rank.

@@

The power spectrum results for inverse covariance weighted fringe-rate filtered data is shown in Figure \ref{fig:toy_sigloss7}. As expected, there is a much larger amount of signal loss for this time-averaged dataset. Additionally, because of the steepened eigenspectrum, most of our power spectrum information is coming from only the last few modes and as a result, it is a noisier estimate. This is evident by noticing that the green curve in Figure \ref{fig:toy_sigloss7} fails to trace the shape of the unweighted EoR power spectrum.

Using our toy model, we have seen that a sensitivity-driven analysis technique like fringe-rate filtering has trade-offs of signal loss and noisier estimates when using data-estimated covariance matrices. Longer integrations increase sensitivity but reduce the number of independent samples, resulting in weakly characterized, steep eigenspectra that can overfit noise greatly. \cc{help me fix explanation}

@@

As a first test, we know that our simulated EoR signal should have a covariance matrix that mimics the identity matrix, with its variance encoded along the diagonal. If we model $\textbf{C}_{EoR}$ as such, instead of computing it based on the dataset itself, and add it to $\hat{\textbf{C}}_{FG}$ (defined as $\hat{\textbf{C}}_{FG} = \langle\textbf{x}_{FG}\textbf{x}_{FG}^{\dagger}\rangle$) to obtain a final $\hat{\textbf{C}}$ to use in weighting, we see that there is negligible signal loss (Figure \ref{fig:toy_sigloss8}, upper left). This is because by modeling $\textbf{C}_{EoR}$, we avoid over-fitting small fluctuations in the data that our model doesn't know about (but an empirically derived $\hat{\textbf{C}}$ would). This is evident when comparing the green and red curves in Figure \ref{fig:toy_sigloss14}. We don't see perfect EoR recovery though, as the green and red curves vary slightly in Figure \ref{fig:toy_sigloss8}. This deviation is a consequence of the difference between the true and modeled covariance matrices.

@@

\begin{figure}
	\centering
	\includegraphics[trim={0.3cm 0.3cm 0.3cm 0.3cm},clip,width=\columnwidth]{plots/toy_sigloss14.png}
	\caption{We compare the eigenspectrum of an empirically calculated $\hat{\textbf{C}}$ (green) to that of four alternate weighting options, including modeling the covariance matrix of EoR (red), regularizing $\hat{\textbf{C}}$ by adding an identity matrix to it (blue), using only the first three eigenmodes of $\hat{\textbf{C}}$ (yellow), and multiplying an identity matrix with $\hat{\textbf{C}}$ (magenta). All eigenspectra (except the green) are relatively flat and don't exhibit signal loss.}
	\label{fig:toy_sigloss14}
\end{figure}

The second panel (top right) in Figure \ref{fig:toy_sigloss8} uses a regularization method of setting $\hat{\textbf{C}} = \hat{\textbf{C}}_{ij} + \gamma\textbf{I}_{ij}$, where $i$$=$$j$$=$$20$ (number of frequencies) and $\gamma = 5$ (an arbitrary strength of $\textbf{I}$ for the purpose of this toy model). By adding the identity matrix, element-wise, we are weighting the diagonal elements of the matrix more heavily than those off-diagonal, thereby flattening out its eigenspectrum. If all modes are given similar weights, we avoid down-weighting artificial modes more than others.

The third panel (bottom left) in Figure \ref{fig:toy_sigloss8} flattens out the eigenspectrum of $\hat{\textbf{C}}$ a different way - by only using the first three eigenmodes. This means that all but the three highest weights will be replaced with $1$'s (equal weights). Again, flattening the eigenspectrum results in negligible signal loss. However, we do not perfectly recover the shape of EoR because we lost information when projecting out certain modes.

The last regularization scheme we are highlighting here is setting $\hat{\textbf{C}} = \hat{\textbf{C}}_{ij}\textbf{I}_{ij}$, or inverse variance weighting (keeping only the diagonal elements of $\hat{\textbf{C}}$). In the bottom right panel of Figure \ref{fig:toy_sigloss8}, we see that this method does a poor job down-weighting foregrounds. For this toy model, our foregrounds are spread out in frequency and therefore have non-negligible frequency-frequency correlations. Multiplying by the identity, element-wise, results in a diagonal matrix, meaning we are only left with correlation information between the same two frequencies. Because we disregard information from all other frequency combinations in this case, we do a poor job suppressing the foreground. But because we flattened the eigenspectrum, we also avoid signal loss.

@@

Our second major $21$ cm power spectrum theme is error estimation, as we desire robust methods for determining accurate confidence intervals for our measurements. Two popular ways of estimating errors on a power spectrum measurement are calculating the variance of a dataset, and computing a theoretical error estimate based on an instrument's system temperature and observational parameters. In a perfect world, both methods would match up. However, in practice the two don't always agree due to a number of factors, including time, frequency, and antenna-dependent noise and non-uniform weightings. Since modeling errors for a complex instrument is only approximate, we choose to compute error bars on our measurements that have been derived from its inherent variance.

A common technique used to estimate the error in a measurement is bootstrapping. Bootstrapping uses sampling with replacement to estimate a posterior distribution. For example, power-spectral measurements of $21$ cm data can be made along many axes, including time and baselines. Averaging each sample set gives us a realization of the power spectrum answer. Realizations are correlated with each other to a degree set by the fraction of sampled points that are held in common between them. Hence, through the process of resampling and averaging along these axes, we can estimate error bars for our results which represent the underlying distribution of power spectra values that are allowed by our measurements.

Again, we focus on toy models to highlight traps that one can fall into when bootstrapping power spectra. Suppose we have a Gaussian random signal dataset of length $N=1000$ and unity variance. We are interested in the average of the dataset, and predict that the error on the mean should obey $1/\sqrt{N}$, where $N$ is the number of samples.

@@

We choose to use $\hat{\textbf{C}} = \hat{\textbf{C}}_{ij}\textbf{I}_{ij}$, or inverse variance weighting, for the remainder of this paper, which we will denote as $\hat{\textbf{C}}_{eff}$. We choose this weighting scheme due to its simplicity in practice and because its effectiveness is comparable to other weighting schemes that produce minimal signal loss.

>>>>

To build our intuition into how a particular choice of weighting, namely inverse covariance weighting, can lead to signal loss, we use a toy model. We construct a simple dataset that contains 2-dimensional data (representing visibility data with $100$ time integrations and $20$ frequency channels). This model represents realistic dimensions of about an hour of PAPER data which might be used for a power spectrum analysis. For PAPER-64 we use $\sim$$8$ hours of data for our analysis (with channel widths of $0.5$ MHz and integration times of $43$ seconds), but here we scale it down for this toy model with no loss of generality.

We create mock visibilities, $\textbf{x}$, and assume a non-tracking, drift scan observation. Hence, flat spectrum sources (away from zenith) oscillate in time and frequency in our measurements. We therefore form a mock bright foreground signal, $\textbf{x}_{FG}$, as a complex sinusoid that varies smoothly in time and frequency, a simplistic but realistic representation of a single bright source. We also create a mock EoR signal, $\textbf{x}_{EoR}$, as a complex, Gaussian random signal. A more realistic EoR signal would have a sloped power spectrum in $\hat{\textbf{p}}$(k) (instead of flat, as in the case of white noise), which could be simulated by introducing frequency correlations into the mock EoR signal. However, in this paper we treat all $k$'s separately, so a simplistic white noise approximation can be used. Our combined data vector is then $\textbf{x} = \textbf{x}_{FG} + \textbf{x}_{EoR}$, to which we apply different weighting schemes. The three data vectors are shown in Figure \ref{fig:toy_sigloss1}.

@@

One choice for the weighting matrix $\textbf{R}$ is an inverse covariance matrix. This type of weighting is attractive for power spectrum analyses because it is an aggressive way to down-weight foregrounds and yields the smallest possible error bars on a measurement (\citealt{tegmark_et_al1997a}; \citealt{bond_et_al1998}). The covariance matrix $\textbf{C}$, which in our case describes covariances between frequency channels, can be estimated in a variety of ways. Given perfect foreground, instrumental, and EoR models, we could form $\textbf{C}$ in a way that accurately describes our measured data. However, if our foreground model is flawed, for example, our estimate of $\textbf{C}$ would not be successful at down-weighting them in the data.

@@

\noindent assuming $\langle\textbf{x}\rangle = 0$ (a reasonable assumption since fringes average to $0$ over a sufficient amount of time), where $\langle \rangle$ denotes a finite average over time $t$. The inverse covariance matrix is therefore $\textbf{R} = \hat{\textbf{C}}^{-1}$.

First, we compute the power spectrum of $\textbf{x}$ using OQE formalism and a weighting matrix of $\hat{\textbf{C}}^{-1}$. The result is shown in green in the left plot of Figure \ref{fig:toy_sigloss3} (the right plot shows an alternate weighting approach and will be discussed further below). Also plotted in the figure are the unweighted ($\textbf{R} = \textbf{I}$) power spectrum of $\textbf{x}_{FG}$ (blue) and $\textbf{x}_{EoR}$ (red).

@@

As shown, our inverse covariance-weighted result successfully suppresses foregrounds. It is also evident that our result fails to recover the EoR signal --- it exhibits the correct shape, but the amplitude level is slightly low. This is evidence of signal loss. In order to understand the behavior of this result, we can closely study our covariance matrix, $\hat{\textbf{C}}$. %, which is shown in Figure \ref{fig:toy_sigloss12}.

If $\textbf{C}$ is computed from the data itself, it carries the risk of over-fitting information in the data and introducing a multiplicative bias (per $k$) to estimates of the signal. For a mathematical derivation of signal loss arising from a data-estimated covariance matrix, see Appendix \ref{sec:sigloss_appendix}. Here we will describe the origin of this signal loss intuitively.

@@

In general, the strongest modes of $\hat{\textbf{C}}$ (highest eigenvalues) are expected to be dominated by bright foregrounds --- the most prominent shapes in the data. If bright foregrounds dominate a dataset, these highest eigenvalue modes would exist in both $\textbf{C}$ and $\hat{\textbf{C}}$. This is demonstrated in the toy model by the successful suppression of the foreground mode, where our estimated covariance matrix identifies the sinusoid shape and assigns it the highest eigenvalue (the peak in Figure \ref{fig:toy_sigloss2}). In Figure \ref{fig:toy_sigloss12} we show the covariances of our toy model datasets along with inverse covariance weighted data. The foreground sinusoid is clearly visible in $\hat{\textbf{C}}_{FG}$ but effectively absent from $\hat{\textbf{C}}^{-1}_{FG}\textbf{x}_{FG}$.

@@

Using what we've learned about the eigenspectrum, we can tweak it in a simple way to suppress foregrounds and yield zero signal loss. Recall that our toy model foreground is a sinusoid that can be perfectly described by a single eigenmode. We can project out the zeroth eigenmode (i.e. zero out all eigenmodes except the first one), thereby down-weighting the foregrounds perfectly and nothing else. Hence, we are not using information from all the weaker eigenmodes which carry the risk of describing fluctuations that may not exist in the true covariance. Altering $\hat{\textbf{C}}$ as such is one example of a regularization method, in which we are changing $\hat{\textbf{C}}$ in a way that flattens its eigenspectrum, making it more similar to that of $\textbf{I}$. The resulting power spectrum estimate for this case is shown in the right plot of Figure \ref{fig:toy_sigloss3}. In this case we recover EoR, demonstrating that if we can model our foregrounds perfectly (and disentangle them from other modes), we can down-weight them without signal loss. There are several other ways to regularize $\hat{\textbf{C}}$, and we will discuss some in Section \ref{sec:otherweight}.

@@

A fringe-rate filter is an analysis technique designed to maximize sensitivity by integrating in time (\citealt{parsons_et_al2016}). Rather than a traditional box-car average, a time domain filter can be designed to up-weight temporal modes corresponding to the sidereal motion on the sky, while down-weighting modes which are noise-like.

Because fringe-rate filtering is analogous to averaging in time, it comes at the cost of reducing the total number of independent samples in the data. To mimic this filter, we average every four time integrations of our toy model dataset together, yielding $25$ independent samples in time (Figure \ref{fig:toy_sigloss5}, left). We choose these numbers so that the total number of independent samples is similar to the number of frequency channels --- therefore, our matrices will be full rank.\footnote{If instead we average every $20$ samples together, yielding only $5$ independent samples in time, we are left with a rank deficient covariance matrix. In this case... \cc{fill in}}

@@

The power spectrum results for inverse covariance weighted fringe-rate filtered data is shown in Figure \ref{fig:toy_sigloss7}. As expected, there is a much larger amount of signal loss for this time-averaged dataset. Additionally, because of the steepened eigenspectrum, the last few eigenmodes receive more weight relative to the others. Therefore, most of our power spectrum information is coming from only these last few modes and as a result, it is a noisier estimate. This is evident by noticing that the green curve in Figure \ref{fig:toy_sigloss7} fails to trace the shape of the unweighted EoR power spectrum.

Using our toy model, we have seen that a sensitivity-driven analysis technique like fringe-rate filtering has trade-offs of signal loss and noisier estimates when using data-estimated covariance matrices. Longer integrations increase sensitivity but reduce the number of independent samples, resulting in poorly characterized, steep eigenspectra that can overfit noise greatly. We note that a fringe-rate filter does have a range of benefits, many described in \citet{parsons_et_al2016}, so it can still be advantageous to use one despite the trade-offs.

@@

As a first test, we model the covariance matrix of EoR as a proof of concept that if perfect models are known, signal loss can be avoided. We know that our simulated EoR signal should have a covariance matrix that mimics the identity matrix, with its variance encoded along the diagonal. We model $\textbf{C}_{EoR}$ as such, instead of computing it based on $\textbf{x}_{EoR}$ itself. Next, we add $\textbf{C}_{EoR} + \hat{\textbf{C}}_{FG}$ (where $\hat{\textbf{C}}_{FG} = \langle\textbf{x}_{FG}\textbf{x}_{FG}^{\dagger}\rangle$) to obtain a final $\hat{\textbf{C}}$ to use in weighting. In Figure \ref{fig:toy_sigloss8} (upper left), we see that there is negligible signal loss. This is because by modeling $\textbf{C}_{EoR}$, we avoid over-fitting small fluctuations in the data that our model doesn't know about (but an empirically derived $\hat{\textbf{C}}$ would). This is evident when comparing the green and red curves in Figure \ref{fig:toy_sigloss14}. We don't see perfect EoR recovery though, as the green and red curves vary slightly in Figure \ref{fig:toy_sigloss8}. This deviation is a consequence of the difference between the true and modeled covariance matrices. In practice such a weighting option is not feasible, as we do not know $\textbf{C}_{EoR}$.

@@

\begin{figure}
	\centering
	\includegraphics[trim={0.3cm 0.3cm 0.3cm 0.3cm},clip,width=\columnwidth]{plots/toy_sigloss14.png}
	\caption{We compare the eigenspectrum of an empirically calculated $\hat{\textbf{C}}$ (green) to that of four alternate weighting options, including modeling the covariance matrix of EoR (red), regularizing $\hat{\textbf{C}}$ by adding an identity matrix to it (blue), using only the first three eigenmodes of $\hat{\textbf{C}}$ (yellow), and multiplying an identity matrix with $\hat{\textbf{C}}$ (magenta). All eigenspectra (except the green) are relatively flat and don't exhibit signal loss. All were computed for the fringe-rate filtered toy model case.}
	\label{fig:toy_sigloss14}
\end{figure}

The second panel (top right) in Figure \ref{fig:toy_sigloss8} uses a regularization method of setting $\hat{\textbf{C}} = \hat{\textbf{C}} + \gamma\textbf{I}$, where $i$$=$$j$$=$$20$ (number of frequencies) and $\gamma = 5$ (an arbitrary strength of $\textbf{I}$ for the purpose of this toy model). By adding the identity matrix, element-wise, we are weighting the diagonal elements of the matrix more heavily than those off-diagonal, thereby flattening out its eigenspectrum. If all modes are given similar weights, we avoid down-weighting artificial modes more than others.

The third panel (bottom left) in Figure \ref{fig:toy_sigloss8} flattens out the eigenspectrum of $\hat{\textbf{C}}$ a different way - by only using the first three eigenmodes. Recalling that the foregrounds can be described entirely by the first eigenmode, this method intentionally projects out modes that are EoR-dominated by replacing all but the three highest weights in the eigenspectrum with $1$'s (equal weights). Again, flattening the eigenspectrum results in negligible signal loss. However, we do not perfectly recover the shape of EoR because we lost information when projecting out certain modes.

The last regularization scheme we are highlighting here is setting $\hat{\textbf{C}} = \hat{\textbf{C}} \circ \textbf{I}$ (element-wise multiplication), or inverse variance weighting (keeping only the diagonal elements of $\hat{\textbf{C}}$). In the bottom right panel of Figure \ref{fig:toy_sigloss8}, we see that this method does a poor job down-weighting foregrounds. For this toy model, our foregrounds are spread out in frequency and therefore have non-negligible frequency-frequency correlations. Multiplying by the identity, element-wise, results in a diagonal matrix, meaning we are only left with correlation information between the same two frequencies. Because we disregard information from all other frequency combinations in this case, we do a poor job suppressing the foreground. But because we flattened the eigenspectrum, we also avoid signal loss.

@@

Our second major $21$ cm power spectrum theme is error estimation, as we desire robust methods for determining accurate confidence intervals for our measurements. Two popular ways of empirically calculating errors on a power spectrum measurement are calculating the variance of a dataset, and computing a theoretical error estimate based on an instrument's system temperature and observational parameters. In a perfect world, both methods would match up. However, in practice the two don't always agree due to a number of factors, including time, frequency, and antenna-dependent noise and non-uniform weightings. For PAPER's analysis, we choose to compute error bars on our measurements that have been derived from its inherent variance.

A common technique used to estimate the error in a measurement is bootstrapping. Bootstrapping uses sampling with replacement to estimate a posterior distribution. For example, power-spectral measurements of $21$ cm data can be made from different samples of data (i.e. data can be sliced along time, baselines, etc. and a power spectrum measurement can be made for each). Averaging each sample set gives us a realization of the power spectrum answer. Realizations are correlated with each other to a degree set by the fraction of sampled points that are held in common between them. Hence, through the process of resampling and averaging along these axes, we can estimate error bars for our results which represent the underlying distribution of power spectra values that are allowed by our measurements.

Again, we focus on toy models to highlight pitfalls that one can fall into when bootstrapping power spectra. Suppose we have a Gaussian random signal dataset of length $N=1000$ and unity variance. We are interested in the average of the dataset, and predict that the error on the mean should obey $1/\sqrt{N}$, where $N$ is the number of samples.

@@

We choose to use $\hat{\textbf{C}} = \hat{\textbf{C}} \circ \textbf{I}$, or inverse variance weighting, for the remainder of this paper, which we will denote as $\hat{\textbf{C}}_{eff}$. We choose this weighting scheme due to its simplicity in practice and because its effectiveness is comparable to other weighting schemes that produce minimal signal loss.
